Historic collections as a tool for assessing the global pollination crisis

### 2018-04-06 ms.Rmd
Main text ready

---
title: Museum collections as baseline for assessing the global pollinator crisis
author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
csl: data/philb_style.xml
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
bibliography: data/references.xml
---

@@

*Affiliations:*
1 Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain   
2 Centre for Biodiversity and Biosecurity, School of Biological Sciences, The University of Auckland, Auckland, New Zealand
3 Landcare Research, Auckland, New Zealand
4 Andrena Soluciones  
*Correspondence: nacho.bartomeus@gmail.com

@@

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data are scarce and often geographically and taxonomically biased. These biases limit robust inference about any potential pollinator crisis. Analysis of historical time series data is the most direct approach for assessing species declines, but long-term monitoring programs targeting pollinators are limited. However, non-structured and often opportunistic specimen collection data are more frequently available. In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits). Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, morphology and pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent. Furthermore, we identify the lack of collection record data digitization as a key bottleneck impeding progress in this field. We highlight opportunities and limitations to progress assessment of pollinator declines globally. Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis. To progress this field, data must be made feely accessible to researchers and appropriate analytical treatment is imperative.

@@

One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Winfree2011]. Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017]. However, bees are diverse, with more than 20,000 species currently described worldwide [@Archer2014], and often require expert taxonomists for identification. Furthermore, the uneven distribution of researchers has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].

@@

At a global scale, current evidence of pollinator declines is highly limited with most data restricted to the US and Europe. It is unsurprising that studies on pollinator declines are biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance. For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collection data detected strong richness declines for bees, hoverflies and flowering plants [@Biesmeijer2006]. However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for natural areas. Further exploration of this dataset revealed that for declining pollinator taxa, the trend has attenuated in recent decades [@Carvalheiro2013].

In the Netherlands, museum data have revealed simultaneous plant and pollinator declines [@Scheper2014]. Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction. Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable. For example, regional estimates for species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012]. This is a pattern also detected in the UK [@Ollerton2014]. However, these regional US findings are in stark contrast with the widespread local extinctions reported by [@Burkle2013], who compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants. It is important to note that [@Burkle2013] focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (@Collado2018). In any case, there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.

As with bees, studies on species richness changes for other pollinator taxa are both scarce and geographically restricted. Generally speaking, reported declines for bumblebees are the most severe of all pollinator taxa. For example, declines of up to 18% in local bumblebee richness have been reported for Belgium and the Netherlands [@Carvalheiro2013]. In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011]. In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a]. For butterflies, the only evidence of richness declines comes from Europe. Butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although declines in Great Britain have been less severe [@Carvalheiro2013]. In Belgium, another study [@Maes2001] found that richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to [@Carvalheiro2013] (1950-69 vs. 1970-80 and 1970-89 vs. 1990-2009). In parts of Germany, up to 70% declines in local butterfly richness have been reported [@Wenzel2006]. Compared with other insect pollinator taxa, there are very few studies on hoverfly species richness changes, which are all restricted to Europe. In Belgium, Great Britain and the Netherlands, hoverfly richness changes have been modest [@Carvalheiro2013]. In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer2006]. Furthermore, directionality (richness increase or decrease) varies depending on the time period assessed. For example, hoverfly richness decreased in Belgium by approximately 6% from 1950-69 to 1970-80, but increased by approximately 10% between 1970-89 and 1990-2009 [@Carvalheiro2013].

Figure one shows (a) bee species richness weighted by country area; (b) bee species richness changes; (c) syrphid species richness changes; and (d) butterfly species richness changes overtime from known studies, to highlight the little coverage that currently exists globally. See raw data (Sup Mat 1). For other parts of the world, there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; @Morales2013) [South Africa: @Pauw, maybe with plants… . However, Peter Keil flies? Moths? Butterflies!]() Outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records.

@@

#read in data
declines <- read.csv(file = "data/richness_declines.V2.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
#nothing outside EU- USA. Spain and NZ can be added.

@@

####################################################
#make syrphid map
####################################################

declines.syrphid <- as.data.frame(filter(declines.ave, taxa == "syrphid")) %>% droplevels()
declines.syrphid.2 <- declines.syrphid %>% 
    group_by(country) %>% 
    summarise(average = mean(average),
              median = median(median),
              max = max(max),
              min = min(min))

@@

####################################################
#make bee map
####################################################

declines.bee <- as.data.frame(filter(declines.ave, taxa == "bee")) %>% droplevels()
declines.bee.2 <- declines.bee %>% 
    group_by(country) %>% 
    summarise(average = mean(average),
              median = median(median),
              max = max(max),
              min = min(min))

####################################################
#make butterfly map
####################################################

declines.butterfly <- as.data.frame(filter(declines.ave, taxa == "butterfly")) %>% droplevels()
declines.butterfly.2 <- declines.butterfly %>% 
    group_by(country) %>% 
    summarise(average = mean(average),
              median = median(median),
              max = max(max),
              min = min(min))

@@

Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size. This accounts for the species-area relationship. Warmer colours indicate higher bee diversity. Note that some African countries may have incomplete listed faunas and that Alaska is included with US values. Changes in (b) syrphid (c) bee and (d) butterfly richness within the last 100 years. Warmer colours indicate steeper declines. Countries without data are coloured in white.

@@

Estimates of pollinator declines are lacking for most countries worldwide (Figure 1). The use of historical museum collection data may be the most effective tool for filling these gaps. The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens, and at the same time, often ensure higher quality taxonomic identification. Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization. Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@Meier2004; @Huber1998; @Thomson2018], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@Hart2016]. Undertaking this process for tens or hundreds of thousands of museum collection specimens can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution photos of specimens and associated labels can be uploaded to the internet, where the task of image transcription can be distributed across hundreds or thousands of volunteers (e.g., https://www.zooniverse.org/). In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage]. However, achieving this requires adequate funding [@Wiggins1991].

@@

As we have shown, there is a paucity of countries for which historical data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines. While ideally one would aim to digitize all museum collection records, this is unlikely in the near future, predominantly due to funding constraints. Here we show how researchers can optimize the use of museum collection data to assess long-term pollinator population changes.

GBIF (https://www.gbif.org/) is a central repository for global species occurrence data. Much of these data come from museums, private collections and government research institutes, but several other sources are also integrated. In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability can be checked for the region of interest. We show the number of modern and historic bee records currently available for different countries (Figure 2). Although values are high on both axes for countries such as the UK and the US (these data can potentially be analysed without further data collection effort), most countries fall short in one or both axes. Hence, for countries with many recent records, researchers should prioritize the digitalization of old material before embarking on data analyses (e.g. Switzerland,... ). In contrast, for countries with large numbers of historic records, re-surveys are required before analysis [e.g., @Cameron2011]. Interestingly, we show that greater than 100 [check]() countries have less than 1,000 records, making them poor candidates for analysing long-term pollinator population trends. However, countries like X or X… It is also important to note that historical records are not always vouchered in local museums (i.e., many European and US museums contain large collections of pollinators from other countries). Aside from bees, similar exploratory analyses can easily be conducted for other taxa. [expand on this?]

@@

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
#abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

@@

#table?
lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) < 7.5),]
# poland switcherland need old records! > 10000 vs 300
# sri lanka and french guayana, 
# nicaragua, venezuela, Ecuador..>2000 vs >500.
# Colombia & Panama, zimbawe > 3000 vs >1500
nrow(lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),])
lhf[which(lhf$country == "ES" | lhf$country == "NZ"),]
col <- ifelse(lhf$country == "ES" | lhf$country == "NZ", "red", "black")
col <- ifelse(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5, col, "grey")
plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col)
temp <- subset(lhf, country %in% c("ES", "NZ"))
text(x = temp$log_post1980+1, y = temp$log_pre1980+1 , labels = temp$country)
#36 countries!
#exp(7.5) #> 1800 records in both axes!
```

@@

```{r map rgbif, echo=FALSE, message=FALSE, warning=FALSE}
countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
load("data/gbif.RData")
#3) split by year
library(vegan)
post1980 <- c()
pre1980 <- c()
n <- c()
for(i in 1:length(countries)){
    dat <- occ_list[[i]]
    nrow(dat) 
    dat1 <- dat[-which(is.na(dat$name)),]
    nrow(dat1) # 
    dat1$period <- ifelse(dat1$year < 1980, "pre1980", "post1980")
    table(dat1$period) # :(
#4) rar richness
    comm <- dcast(dat1, period ~ name, fun.aggregate = length, value.var = "name")
    rownames(comm) <- comm$period
    comm <- comm[,-1]
    temp <- rarefy(comm, sample = round(min(table(dat1$period))*0.75))
    #str(temp)
    post1980[i] <- temp[1]
    pre1980[i] <- temp[2]
    n[i] <- round(min(table(dat1$period))*0.75)
}
#5) t test!
data <- data.frame(countries = as.character(countries), pre1980 = pre1980, post1980 = post1980, n = n)
head(data)
#maybe delete n < 500?
data2 <- subset(data, n > 500)

t.test(data2$pre1980, data2$post1980, paired = TRUE)

boxplot(data2$pre1980, data2$post1980, names = c("pre1980", "post1980"), ylab = "rarefied species richness", 
            las = 1, border = "light grey")
points(x = c(rep(1,nrow(data2)), rep (2,nrow(data2))), y = c(data2$pre1980, data2$post1980), pch = 19)
for(i in 1:nrow(data2)){
    lines(x = 1:2, y = c(data2$pre1980[i], data2$post1980[i]))
}

@@

#average richness change
map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette="topo",
                       mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))

@@

Figure 2. a) The number of bee occurrences before 1980 and after 1980 in GBIF. The upper right quadrat contains well covered countries and the upper left quadrat contains countries that require resurveys of old sites. The lower right quadrat contains countries that require museum record digitization and the lower left quadrat contains countries with very little available data. We highlight a few representative examples; b) a boxplot showing the rarefied number of species collected before and after 1980 and; c) the same data plotted on a global map. This data are likely to contain strong undetected biases, but indicate little change...

Once museum collection datasets are made available, researchers must identify any potential biases. [Talk about the caveats of the above crude analysis!]() We explore this process with two contrasting dataset examples (Spain and New Zealand). In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people. In this case, we contacted the original collectors to define their sampling protocols. We then resurveyed the same sites (20 years after the original surveys) using the same sampling protocols. In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time. For these case studies, we provide annotated R scripts for analysis of different dataset types (e.g., for Spain and New Zealand, as described above). These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyse data for regions where current information on pollinator declines is lacking.

@@

Spain provides an interesting study system because its natural habitats have been transformed extensively by humans over a long time period, but land-use is not as intensive compared with many other European countries. In addition, Spain is a bee diversity hotspot (Figure 1) and maintains a relatively heterogeneous landscape. Spain has already digitalized a large amount of pollinator occurrence data for both historic and recent periods (Figure 2). However, visual inspection of the data revealed clustering around a few localities. Further, historic records did not spatially match recent records, making comparisons difficult. For this dataset, most of the historic records were located around Valladolid and were collected by Enrique Asensio and collaborators. There has been no recent sampling of bees in this area. However, we found that Enrique systematically sampled seven locations and that additional data were available at the “Museo de Historia Nacional” and other minor collections. Digitization of these records, along with a re-survey of the original sampling locations provided an excellent dataset for a before and after comparison of bee communities. Interestingly, these localities have recently experienced substantial agricultural intensification [expand]()

In brief, we cleaned taxonomic names using the taxize package [@Chamberlain2018], checked sampling completeness for both time periods and compared species richness for each site before and after 1980 with a paired t-test. We found that there were a reduced number of species at sites after 1980 (P = 0.058). However, this trend was highly dependent on site identity, as two out of three sites showed no richness declines. However, we found clear patterns in declines among species groups. For example, Andrenids and their parasites showed the strongest declines whereas Halictids increased. This is in accordance with findings elsewhere [e.g. @Bartomeus2013a] (appendix?).

>>>>

---
title: Historic collections as a tool for assessing the global pollination crisis
author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
csl: data/philb_style.xml
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
bibliography: data/references.xml
---

@@

*Affiliations:*
^1^ Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain   
^2^ Centre for Biodiversity and Biosecurity, School of Biological Sciences, The University of Auckland, Auckland, New Zealand  
^3^ Landcare Research, Auckland, New Zealand  
^4^ Andrena Iniciativas y Estudios Medio Ambientales
*Correspondence: nacho.bartomeus@gmail.com

@@

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data are scarce and often geographically and taxonomically biased. These biases limit robust inference about any potential pollinator crisis. 
Non-structured and opportunistic historical specimen collection data provide the only source of historical information which can serve as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums not only provide information on where and when species were collected, but also contain other ecological information such as species interactions and morphological traits. Here, we provide a synthesis of how researchers have used historical data to identify long-term changes in biodiversity, species abundances, morphology and pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent, but we highlight opportunities and limitations to progress the assessment of pollinator declines globally. Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis.

*Keywords*: Museums, biodiversity, global change, bees, syrphid flies, butterflies.

@@

One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017]. Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011]. However, bees are diverse, with more than 20,000 species currently described worldwide, and often require expert taxonomists for identification. Furthermore, the uneven distribution of researchers has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].

@@

At a global scale, current evidence of pollinator declines is highly limited with most data restricted to the US and Europe. It is unsurprising that studies on pollinator declines are biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance. For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collection data detected strong richness declines for bees, hoverflies and flowering plants [@Biesmeijer2006]. In the Netherlands, museum data have also revealed simultaneous plant and pollinator declines [@Scheper2014]. Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction. However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for well-preserved natural areas. Further exploration of this dataset revealed that for declining pollinator taxa, the trend has attenuated in recent decades [@Carvalheiro2013].

Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable. For example, regional estimates for bee species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012]. This is a pattern also detected in the UK, where relatively few regional bee extinctions has been reported [@Ollerton2014]. These regional findings are in stark contrast with the widespread local extinctions reported in local studies. For example, Burkle et al. [@Burkle2013] compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants and reports several local extinctions. However, it is important to note that there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.

Reported declines for bumblebees are the most severe of all pollinator taxa. For example, declines of up to 18% in local bumblebee richness have been reported for Belgium and the Netherlands [@Carvalheiro2013]. In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011]. In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a]. However, studies on species richness changes for other pollinator taxa are both scarce and geographically restricted. For butterflies, the only evidence of richness declines comes from Europe. Butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although declines in Great Britain have been less severe [@Carvalheiro2013]. In Belgium, another study [@Maes2001] found that richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to [@Carvalheiro2013] (1950-69 vs. 1970-80 and 1970-89 vs. 1990-2009). In parts of Germany, up to 70% declines in local butterfly richness have been reported [@Wenzel2006]. Compared with other insect pollinator taxa, there are very few studies on hoverfly species richness changes, which are all restricted to Europe. In Belgium, Great Britain and the Netherlands, hoverfly richness changes have been modest [@Carvalheiro2013]. In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer2006]. Furthermore, directionality (richness increase or decrease) varies depending on the time period assessed. For example, hoverfly richness decreased in Belgium by approximately 6% from 1950-69 to 1970-80, but increased by approximately 10% between 1970-89 and 1990-2009 [@Carvalheiro2013].

For ilustrative purposes, we mapped the findings of this studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1). Despite outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records, there are species-specific examples of historical losses from different parts of the world (e.g., Bombus dalbhomi; @Morales2013).

@@

#read in data
declines <- read.csv(file = "data/richness_declines.V2.csv", 
                     header = TRUE, sep = ",", na.strings=c("","NA"))
#nothing outside EU- USA. Spain and NZ can be added.
#head(declines)

@@

#make bee map
declines.bee <- as.data.frame(filter(declines.ave, taxa == "bee")) %>% droplevels()
declines.bee.2 <- declines.bee %>% 
    group_by(country) %>% 
    summarise(average = mean(average),
              median = median(median),
              max = max(max),
              min = min(min))

#make syrphid map
declines.syrphid <- as.data.frame(filter(declines.ave, taxa == "syrphid")) %>% droplevels()
declines.syrphid.2 <- declines.syrphid %>% 
    group_by(country) %>% 
    summarise(average = mean(average),
              median = median(median),
              max = max(max),
              min = min(min))

@@

#make butterfly map
declines.butterfly <- as.data.frame(filter(declines.ave, taxa == "butterfly")) %>% droplevels()
declines.butterfly.2 <- declines.butterfly %>% 
    group_by(country) %>% 
    summarise(average = mean(average),
              median = median(median),
              max = max(max),
              min = min(min))

@@

Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size. This correction accounts for the species-area relationship. Warmer colours indicate higher bee diversity. Note that some African countries may have incomplete listed faunas and that Alaska is included with US values. Countries with available historical changes in (b) bee, (c) syrphid and (d) butterfly richness within the last 100 years. Warmer colours indicate steeper average declines. Countries without data are coloured in white.

@@

Estimates of pollinator declines are lacking for most countries worldwide (Figure 1). The use of historic  collection data may be the most effective tool for filling these gaps. The core aim of museums is to conserve and curate historic collections. Thus, they serve as a precious repository for specimens, and at the same time, often ensure higher quality taxonomic identification. Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization. Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@Meier2004; @Huber1998; @Thomson2018], (ii) geo-locate the coordinates of collection events (e.g. http://www.geonames.org) and (iii) store the data in a properly curated database [@Hart2016]. Undertaking this process for tens or hundreds of thousands of museum collection specimens can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution photos of specimens and associated labels can be uploaded to the internet, where the task of image transcription can be distributed across hundreds or thousands of volunteers (e.g., https://www.zooniverse.org/). In addition, new algorithms have been created that allow location geo-referencing based on vernacular names (e.g. https://geoparser.io). However, achieving this requires adequate funding [@Wiggins1991].

@@

As we have shown, there is a paucity of countries for which historical data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines. While ideally one would aim to digitize all museum collection records, this is unlikely in the near future, predominantly due to funding constraints. Here we show how researchers can optimize the use of historical collection data to assess long-term pollinator population changes.

GBIF (https://www.gbif.org/) is a central repository for global species occurrence data. Much of these data come from museums, private collections and government research institutes, but several other sources are also integrated. In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability can be checked for the region of interest. Focusing on bee taxa, we show here the number of modern and historic bee records currently available for different countries (Figure 2a, Sup mat 2). Thirty-seven countries count with more than 1800 records in each time period, making these data potentially analyzable without further data collection effort (see Figure 2b and c for an initial exploration). However, a proper analysis of this dataset would need a careful look at the data as we detail later on with two examples (Spain and New Zealand). Most countries fall short in one or both axes. For example, a variety of countries located in different continents such as Switzerland, Sri Lanka, Nicaragua or Zimbawe have a decent number of recent records, but lack historical collections and researchers should prioritize the digitalization of old material before embarking on data analyses. It is also important to note that historical records are not always vouchered in local museums (i.e., many European and US museums contain large collections of pollinators from other countries). In contrast, more than 192 countries have less than 1000 records for both time periods, making them poor candidates for analysing long-term pollinator population trends. Aside from bees, similar exploratory analyses can easily be conducted for other taxa.

@@

```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
#lhf
#head(lhf)
#str(lhf)
lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
write.csv(lhf, "data/SpMat2.csv")
#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
#abline(h = 150000, v = 150000)
#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.

@@

#table?
#lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) < 7.5),]
#lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),]
#nrow(lhf[which(log(lhf$post1980) < 6.9 & log(lhf$pre1980) < 6.9),])
# poland switcherland need old records! > 10000 vs 300
# sri lanka and french guayana, 
# nicaragua, venezuela, Ecuador..>2000 vs >500.
# Colombia & Panama, zimbawe > 3000 vs >1500
#lhf[which(lhf$country == "ES" | lhf$country == "NZ"),]
col <- ifelse(lhf$country == "ES" | lhf$country == "NZ", "red", "black")
col <- ifelse(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5, col, "grey")
pchs <- ifelse(lhf$country == "ES" | lhf$country == "NZ", 19, 1)
plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs)
temp <- subset(lhf, country %in% c("ES", "NZ"))
text(x = temp$log_post1980, y = temp$log_pre1980+2 , labels = temp$country, col = "red")
#36 countries!
#exp(7.5) #> 1800 records in both axes!
```

@@

```{r map rgbif, echo=FALSE, message=FALSE, warning=FALSE}
countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
load("data/gbif.RData")
#3) split by year
library(vegan)
post1980 <- c()
pre1980 <- c()
n <- c()
for(i in 1:length(countries)){
    dat <- occ_list[[i]]
    #nrow(dat) 
    dat2 <- dat[-which(is.na(dat$name)),]
    dat3 <- dat2[which(is.na(as.numeric(dat2$name))),]
    dat4 <- subset(dat3, name != "Apis mellifera")
    #nrow(dat1) 
    dat1 <- dat4
    #unique(dat1$name)
    dat1$period <- ifelse(dat1$year < 1980, "pre1980", "post1980")
    #table(dat1$period) # :(
#4) rar richness
    comm <- dcast(dat1, period ~ name, fun.aggregate = length, value.var = "name")
    rownames(comm) <- comm$period
    comm <- comm[,-1]
    temp <- rarefy(comm, sample = round(min(table(dat1$period))*0.75))
    #str(temp)
    post1980[i] <- temp[1]
    pre1980[i] <- temp[2]
    n[i] <- round(min(table(dat1$period))*0.75)
}
#5) t test!
data <- data.frame(countries = as.character(countries), pre1980 = pre1980, post1980 = post1980, n = n)
#head(data)
#maybe delete n < 500?
data2 <- subset(data, n > 500) #28 countries with enought data.

#t.test(data2$pre1980, data2$post1980, paired = TRUE)

boxplot(log(data2$pre1980), log(data2$post1980), names = c("pre1980", "post1980"), ylab = "log rarefied species richness", 
            las = 1, border = "light grey")
points(x = c(rep(1,nrow(data2)), rep (2,nrow(data2))), y = c(log(data2$pre1980), log(data2$post1980)), pch = 1)
for(i in 1:nrow(data2)){
    lines(x = 1:2, y = c(log(data2$pre1980[i]), log(data2$post1980[i])), 
          col = ifelse((data2$pre1980[i] - data2$post1980[i]) > 1, "red", "blue"))
}

@@

#average richness change
colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
colorss <- (heat.colors(10))
map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
                       mapTitle = "Average bee richness change (%)", catMethod = "pretty", addLegend = F)
do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))

@@

Figure 2. Exploration of available data for bee records showing a) The number of bee occurrences before 1980 and after 1980 in GBIF for each country. The upper right quadrat (records in black) contains well covered countries with New Zealand (NZ) and Spain (ES) marked in red. For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is similar in both time periods (b). A more careful analysis of this data would help complete the map of global declines (c). Note that this data are likely to contain strong undetected biases, as we explore below.

As stated above, once historical collection datasets are made available, researchers must identify any potential biases. We explore this process with two contrasting dataset examples (Spain and New Zealand). In the Spanish dataset, most of the data comes from a few specific locations and was collected by a few specific teams. Hence the geographical coverage is not representative. Even worst, historical and modern collections do not overlap spatially, making any inference imposible to interpret. In this case, contacted the original collectors of the historical data to define their sampling protocols. We then resurveyed the same sites (30 years after the original surveys) using the same sampling protocols. In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time. We complemented gbif data with further museum collections for bees and flies and analyze the regional richness changes through time. For these two case studies, we provide annotated R scripts as examples of analysis for different dataset types (Sup Mat 3). These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyse data for regions where current information on pollinator declines is lacking.

@@

Spain provides an interesting study system because its natural habitats have been transformed extensively by humans over a long time period, but land-use is not as intensive compared with many other European countries. In addition, Spain is a bee diversity hotspot (Figure 1a) and maintains a relatively heterogeneous landscape. Spain has already digitalized a large amount of pollinator occurrence data for both historic and recent periods (Figure 2a). However, visual inspection of the data revealed clustering around a few localities. Further, historic records did not spatially match recent records, making comparisons difficult. For this dataset, most of the historic records were located around Valladolid and were collected by Enrique Asensio and collaborators. There has been no recent sampling of bees in this area. However, we found that Enrique systematically sampled six independent locations and that additional historical data were available at the “Museo de Historia Nacional” and other minor collections. Digitization of these records, along with a re-survey of the original sampling locations provided an excellent dataset for a before and after comparison of bee communities.

In brief, after cleaning taxonomic names for erratas and synonims using the taxize package [@Chamberlain2018], we checked for sampling completeness for both time periods and compared rarefied species richness for each site before and after 1980 with a paired t-test (rarefication at 1000 specimens). We found that there were a reduced number of species at sites after 1980 (mean difference 20.27 species; 95 percent confidence interval: -1.03, 41.58; t = 2.44, df = 5, P = 0.06). However, this trend was highly dependent on site identity, as two out of three sites showed no richness declines. Interestingly, these two localities were the two that has experienced less land use changes (both are natural areas embeded into agro-ecosystemss). In contrast the other 4 localities suffered large urban or agricutural intensification. In addition, species not further recorded in the re-survey are not a random selection of species, but are clustered in a few genera. For example, Andrenids and their parasites (e.g. Nomada) showed the strongest declines whereas Halictids tend to be more stable (Sup mat 4). This pattern of winners and loosers of land use intensification is in accordance with findings elsewhere [e.g. @Bartomeus2013a], indicating that some clades are more sensitive to disturbance than others.
